---
title: "MB5370 - Module 4"
author: "Maria Bologna"
date: "2023-05-04"
output: html_document
---

# WORKSHOP 1 - Reproducible work ####

## Installing packages & dataset

```{r}
install.packages("tidyverse")
library(tidyverse)

#### ctrl + alt + i -> adds a chunk =)

mpg
summary(mpg)
```

## 2.6 Create your first ggplot

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## 2.7 Understand the "grammar of graphics"

### 2.7.1 Graphing templates

```{r}
#ggplot(data = <DATA>) + 
# <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

### 2.7.2 Aesthetic mappings

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) #change point COLOUR by class

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class)) #change point SIZE by class

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class)) #change point TRANSPARENCY by class

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class)) #change point SHAPE by class

## Manually setting properties
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

## 2.8 Troubleshooting

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy))
```

## 2.9 Facet and panel plots

```{r}
# Facet plot with 1 variable
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)

# Facet plot with >1 variable
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl) #facet_grid(), ~ to separate variables
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl) #use a . to not facet in rows or columns (ex. *facet_grid(. ~ cyl)*)
```

## 2.10 Fitting simple lines

```{r}
# Displaying data as POINTS
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) #**geom_point**

# Displaying data as LINES
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) #**geom_smooth**
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) #change line type per variable
 ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

# Plotting multiple geoms
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth() #no variables specified
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth() #display different aesthetics in different layers
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) #select and plot only one subset in a different layer (ex. *(class == "subcompact")*)
```

## 2.11 Transformations and Stats

### 2.11.1 Plotting statistics

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut)) #use **stat_count** for same result
```

### 2.11.2 Overriding defaults in ggplot2

```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
) #override a default stat (count, summary) to identity (variable raw value)
demo
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity") #plot results
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop), group = 1)) #override aesthetic from count to proportion of data set
```

### 2.11.3 Plotting statistical details

```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  ) #show more detail on data set
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop))) #proportion bar chart
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop))) #proportion bar chart
#both bar charts are missing the group=1 input, therefore all classes for the CUT variable are the same size
```

## 2.12 Positional adjustments

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut)) #differentiate x-variable values with bar outline
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) #differentiate x-variable values with bar colors
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity") #alter transparency

ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity") #color the bar outlines with no fill color

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") #make stacked bars the same height maintaining original proportions

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge") #place overlapping objects beside one another

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter") #add some random noise to each point to avoid overlapping in a scatterplot
```

## 2.13 The layered grammar of graphics

```{r}
#ggplot(data = <DATA>) + 
#  <GEOM_FUNCTION>(
#     mapping = aes(<MAPPINGS>),
#     stat = <STAT>, 
#     position = <POSITION>) +
#  <FACET_FUNCTION> #updated template to make a ggplot
```

----------

# WORKSHOP 2 - Using ggplot2 for communication ####

## 3.1 Labels

```{r}
#Adding a title with labs()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size") #*labs(title= )*

#Adding subtitles and caption with labs()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov") #**labs(title= ,subtitle= ,caption= )**

#Replacing axis labels and legend titles
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type") #**labs(x= ,y=, colour= )**
```

## 3.2 Annotations

```{r}
#Making annotations on the graph itself 
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)
```

## 3.3 Scales

```{r}
#Default scales
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

#Customising scales
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous(breaks = seq(0, 10, by = 2)) +
  scale_y_continuous(breaks = seq(15, 40, by = 5)) +
  scale_colour_discrete() #use code from 3.4 chunk to change scale of x and y axis within dataset 
```

## 3.4 Axis ticks

```{r}
#Changing axis ticks from original graph
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))

seq(15,40, by=5) #creating a sequence of numbers

#Deleting axis ticks numerical values
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)
```

## 3.5 Legends and colour schemes

```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) #create a legend

#Changing legend position
base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") #default legend setting

#Hiding legend
base + theme(legend.position = "none") #*legend.position="none" to hide legend after creating one
```

## 3.6 Replacing a scale

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10() #

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) #default colour scale

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1") #custom colour scale

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1") #custom colour scale & different point styles for each data type

# Using predefined colours
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue")) #*scale_colour_manual()

# Using a preset colour package

install.packages('viridis') #package "viridis"
install.packages('hexbin')
library(viridis)
library(hexbin)

df <- tibble( #fake dataset to create a plot 
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + #new geom
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

## 3.7 Themes

```{r}
#Trying different default themes
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw() #theme_bw()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light() #theme_light()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic() #theme_classic()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark() #theme_dark()

#Developing a custom theme
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="pink"),
        axis.text.x=element_text(colour="pink"),
        text=element_text(family="Arial"))
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme(panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="pink"),
        axis.text.x=element_text(colour="pink"),
        text=element_text(family="Arial"))
```

## 3.8 Saving and exporting plots

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")

```

----------

# WORKSHOP 3 - Reproducible work ####

```{r}
#Setting up Github

install.packages("usethis")
library(usethis)
use_git_config(user.name = "thestralcentauro13093", user.email = "bolognamaria21@gmail.com")
```

----------

# WORKSHOP 4 - Data wrangling in R ####

## 5.3 What is a tibble?

```{r}
#library(tidyverse)

#Converting regular data frames in tibbles
iris 
str(iris)
summary(iris)
as_tibble(iris) #*as_tibble()

#Calling new variables
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y)

#Building a dataframe with *data.frame()
#data.frame(c(x = 1:5, y = 1, z = x ^ 2 + y)) --> Error in data.frame(c(x = 1:5, y = 1, z = x^2 + y)) : object 'x' not found
  
#Working with non-syntactic variables
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
tb

#Using tribbles (transposed tibbles)
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5) #*tribble() to help with data entry directly in your script

#Differences with tribbles
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)) #tibbles only print out the first 10 rows of data plus the type of each column variable next to its name

#install.packages("nycflights13")
library(nycflights13)
nycflights13::flights %>% 
  print(n = 10, width = Inf) #print a designated number of rows with *print(n= number of rows, width= number of columns)

#Options with tibbles
#-":" if more than n rows, print only m rows
#-"options(tibble.print_min = Inf)" to always show all rows
#-"options(tibble.width = Inf)" to print all columns, regardless of the screen width
#-Use R’s built-in viewer with "View()" -> "package?tibble" to check complete list of options 
options(tibble.width = Inf)

#Pulling out a single variable from dataframe
df <- tibble(
  x = runif(5),
  y = rnorm(5))

df$x
df[["x"]]#extract by name
df[[1]] #extract by row position
df[[2,2]] #extract by exact position

#Pipes (*%%)
df %>% .$x
df %>% .[["x"]] #use of "." as placeholder when using pipes

#Recalling a partially matching variable
#df <- tibble(xxx = runif(5, y = rnorm(5))
#df$xx #Warning: Unknown or uninitialised column: `xx`.NULL

#Exercise: why are tibbles preferable?
df <- data.frame(abc = 1, xyz = "a")
df
df$x #call by name
df[, "xyz"] #call by exact position
#Tibbles are preferable as they allow to recall single values from a variable in different ways, which can help when dealing with large or unorganised datasets
```

## 5.4 How can I import data?

```{r}
#heights <- read_csv("data/heights.csv") #import data with *read.csv()

read_csv("a,b,c
1,2,3
4,5,6") #supply the data directly from the script

read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2) #to skip over a set number of lines at the top of the dataset -> *skip= n

read_csv("# A comment I want to skip
  x,y,z
  1,2,3", comment = "#") #to drop all lines starting with a certain character -> *comment = "character"

#Solutions for dataframes with nameless columns
read_csv("1,2,3\n4,5,6", col_names = FALSE) #sequentially name columns x1, x2, etc with *col_names= FALSE #shortcut to add a new line: "*\n"

read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z")) #create a vector to be used as column names with *col_names

read_csv("a,b,c\n1,2,.", na = ".") #to specify "no data" cells in your dataframe, use *na= value used to represent missing data
```

## 5.5 Tidying data  using Tidyr

```{r}
library(tidyverse)
```

### 5.5.1 Tidy data

```{r}
table1
table2
table3
table4a
table4b

table1 %>%
  mutate(rate = cases / population * 10000) #compute rate per 10000
table1 %>%
  count(year, wt = cases)#compute rate per year

#Visualizing changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

### 5.5.2 Speading and gathering data tables

```{r}
#Tidying a dataset with numerical values as column names
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases") #to change column names -> *pivot_longer()
table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population") #same as above

#Tidying  dataset with an observation scattered across multiple rows
table2 %>%
    pivot_wider(names_from = type, values_from = count) #to unify observations of the same variable from on different rows -> *pivot_wider()
```

### 5.5.3 Separating and uniting data tables

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population")) #to divide columns that contain 2 or more variables at the same time -> *separate(original_column_name, into = c(vector with new column names))

table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/") #separate variables with a specific character (e.g., /)

table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE) #convert variables listed as characters to better data types -> *convert = TRUE

table3 %>% 
  separate(year, into = c("century", "year"), sep = 2) #split a numerical value in 2+columns after a set number f digits -> *sep= x

table5 %>% 
  unite(new, century, year, sep = "") #unite 2+ columns into one -> *unite(new, x, y, sep="")
```

### 5.5.4 Handling missing values

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)) #missing values as NA or simply absent

stocks %>% 
  pivot_wider(names_from = year, values_from = return) #make explicit the implicit missing value

stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(
    cols = c(`2015`, `2016`), 
    names_to = "year", 
    values_to = "return", 
    values_drop_na = TRUE) #turn implicit missing values into a valid row of data in dataframe -> *values_drop_na= TRUE

stocks %>% 
  complete(year, qtr) #take a set of columns from original dataset, plus fill in explicit NA values where needed -> *complete(x,y)

treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4)
treatment

treatment %>% 
  fill(person) #replace missing values in a dataset with the most recent non-missing value -> *fill(column_with_NAs)
```

## 5.7 Pipes (%%) for more readable flows

```{r}
#Telling a story with R

#Little bunny Foo Foo_Went hopping through the forest_Scooping up the field mice_And bopping them on the head

#library(magrittr)

# foo_foo <- little_bunny()

#1. Create a function for each key verb: hop(), scoop(), bop()
# foo_foo_1 <- hop(foo_foo, through = forest)
# foo_foo_2 <- scoop(foo_foo_1, up = field_mice)
# foo_foo_3 <- bop(foo_foo_2, on = head) #step-by-step construction of each function

#2. Overwrite original objects instead of creating multiple intermediate ones
# foo_foo <- hop(foo_foo, through = forest)
# foo_foo <- scoop(foo_foo, up = field_mice)
# foo_foo <- bop(foo_foo, on = head)

#3. String functions together
# bop(
# scoop(
# hop(foo_foo, through = forest),
# up = field_mice
# ), 
# on = head
# )

#4. Use a pipe
# foo_foo %>%
# hop(through = forest) %>%
# scoop(up = field_mice) %>%
# bop(on = head)

#Classes of functions this pipe won't work for:

#1. Functions already using the current environment
assign("x", 10) #assign() to create a new variable with the iven name in the current environment

"x" %>% assign(100)
x

env <- environment()
"x" %>% assign(100, envir = env) #specify the environment for the %% to work

#2. Functions using lazy evaluation
tryCatch(stop("!"), error = function(e) "An error")

stop("!") %>% 
tryCatch(error = function(e) "An error")
```

----------

# WORKSHOP 5 - Spatial data in R

## 6.4 Installing the spatial R packages

```{r}
install.packages("sf") 
install.packages("terra")
install.packages("tmap")

library(tidyverse)
library(sf) 
library (terra) 
library(tmap)
```

## 6.6 Loading the spatial dataset

```{r}
library(readr)
dat <- read_csv("E:/Module04_NEW/data-for-course/copepods_raw.csv")
dat
```

## 6.7 Data exploration

### 6.7.1 check coordinates

```{r}
library(ggplot2)
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()

ggplot(dat, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()
```

## 6.8 Getting going with maps

```{r}
sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), 
                 crs = 4326)
summary(sdat)
```

## 6.9 Coordinate reference system

```{r}
crs4326 <- st_crs(4326)
crs4326
crs4326$Name

crs4326$wkt
```

## 6.10 Feature collection (points)

```{r}
sdat
```

## 6.11 Cartography

```{r}
plot(sdat["richness_raw"])

plot(sdat)
```

## 6.12 Thematic maps

```{r}
tm1 <- tm_shape(sdat) + 
  tm_dots(col = "richness_raw")

tmap_save(tm1, filename = "Richness-map.png", 
          width = 600, height = 600)

```

## 6.13 Mapping spatial polygons as layers

### 6.13.1 Loading shapefiles

```{r}
aus <- st_read("E:/Module04_NEW/data-for-course/spatial-data/Aussie/Aussie.shp")

shelf <- st_read("E:/Module04_NEW/data-for-course/spatial-data/aus_shelf/aus_shelf.shp")

aus
```

### 6.13.2 Mapping your polygons

```{r}
tm_shape(shelf) + 
  tm_polygons()

tm_shape(shelf, bbox = sdat) + 
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() + 
  tm_shape(sdat) + 
  tm_dots()
```

## 6.14 Exporting your map

```{r}
plot(sdat["route"])

plot(sdat)

tm1 <- tm_shape(sdat) + 
  tm_dots(col = "route")

vignette('tmap-getstarted')
tmap_mode("view")+
tm_shape(shelf, bbox = sdat) + 
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() +
   tm_shape(sdat) + 
  tm_dots(col="route", palette ="Set3")+
tm_tiles("Stamen.TonerLabels")+
tm_basemap("Stamen.Watercolor")
```


