---
title: "WORKSHOP 1 & 2"
author: "Maria Bologna"
date: '2023-05-20'
output: html_document
---

# WORKSHOP 1 - Reproducible work ####

## Installing packages & dataset

```{r, echo=T, results='hide', fig.show='hide'}
#install.packages("tidyverse")
library(tidyverse)

#### ctrl + alt + i -> adds a chunk =)

mpg
summary(mpg)
```

## 2.6 Create your first ggplot

```{r, echo=T, results='hide', fig.show='hide'}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## 2.7 Understand the "grammar of graphics"

### 2.7.1 Graphing templates

```{r, echo=T, results='hide', fig.show='hide'}
#ggplot(data = <DATA>) + 
# <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

### 2.7.2 Aesthetic mappings

```{r, echo=T, results='hide', fig.show='hide'}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) #change point COLOUR by class

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class)) #change point SIZE by class

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class)) #change point TRANSPARENCY by class

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class)) #change point SHAPE by class

## Manually setting properties
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

## 2.8 Troubleshooting

```{r, echo=T, results='hide', fig.show='hide'}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy))
```

## 2.9 Facet and panel plots

```{r, echo=T, results='hide', fig.show='hide'}
# Facet plot with 1 variable
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)

# Facet plot with >1 variable
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl) #facet_grid(), ~ to separate variables
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl) #use a . to not facet in rows or columns (ex. *facet_grid(. ~ cyl)*)
```

## 2.10 Fitting simple lines

```{r, echo=T, results='hide', fig.show='hide'}
# Displaying data as POINTS
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) #**geom_point**

# Displaying data as LINES
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) #**geom_smooth**
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) #change line type per variable
 ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

# Plotting multiple geoms
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth() #no variables specified
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth() #display different aesthetics in different layers
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) #select and plot only one subset in a different layer (ex. *(class == "subcompact")*)
```

## 2.11 Transformations and Stats

### 2.11.1 Plotting statistics

```{r, echo=T, results='hide', fig.show='hide'}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut)) #use **stat_count** for same result
```

### 2.11.2 Overriding defaults in ggplot2

```{r, echo=T, results='hide', fig.show='hide'}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
) #override a default stat (count, summary) to identity (variable raw value)
demo
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity") #plot results
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop), group = 1)) #override aesthetic from count to proportion of data set
```

### 2.11.3 Plotting statistical details

```{r, echo=T, results='hide', fig.show='hide'}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  ) #show more detail on data set
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop))) #proportion bar chart
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop))) #proportion bar chart
#both bar charts are missing the group=1 input, therefore all classes for the CUT variable are the same size
```

## 2.12 Positional adjustments

```{r, echo=T, results='hide', fig.show='hide'}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut)) #differentiate x-variable values with bar outline
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) #differentiate x-variable values with bar colors
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity") #alter transparency

ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity") #color the bar outlines with no fill color

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") #make stacked bars the same height maintaining original proportions

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge") #place overlapping objects beside one another

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter") #add some random noise to each point to avoid overlapping in a scatterplot
```

## 2.13 The layered grammar of graphics

```{r, echo=T, results='hide', fig.show='hide'}
#ggplot(data = <DATA>) + 
#  <GEOM_FUNCTION>(
#     mapping = aes(<MAPPINGS>),
#     stat = <STAT>, 
#     position = <POSITION>) +
#  <FACET_FUNCTION> #updated template to make a ggplot
```

----------

# WORKSHOP 2 - Using ggplot2 for communication ####

## 3.1 Labels

```{r, echo=T, results='hide', fig.show='hide'}
#Adding a title with labs()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size") #*labs(title= )*

#Adding subtitles and caption with labs()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov") #**labs(title= ,subtitle= ,caption= )**

#Replacing axis labels and legend titles
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type") #**labs(x= ,y=, colour= )**
```

## 3.2 Annotations

```{r, echo=T, results='hide', fig.show='hide'}
#Making annotations on the graph itself 
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)
```

## 3.3 Scales

```{r, echo=T, results='hide', fig.show='hide'}
#Default scales
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

#Customising scales
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous(breaks = seq(0, 10, by = 2)) +
  scale_y_continuous(breaks = seq(15, 40, by = 5)) +
  scale_colour_discrete() #use code from 3.4 chunk to change scale of x and y axis within dataset 
```

## 3.4 Axis ticks

```{r, echo=T, results='hide', fig.show='hide'}
#Changing axis ticks from original graph
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))

seq(15,40, by=5) #creating a sequence of numbers

#Deleting axis ticks numerical values
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)
```

## 3.5 Legends and colour schemes

```{r, echo=T, results='hide', fig.show='hide'}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) #create a legend

#Changing legend position
base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") #default legend setting

#Hiding legend
base + theme(legend.position = "none") #*legend.position="none" to hide legend after creating one
```

## 3.6 Replacing a scale

```{r, echo=T, results='hide', fig.show='hide'}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10() #

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) #default colour scale

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1") #custom colour scale

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1") #custom colour scale & different point styles for each data type

# Using predefined colours
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue")) #*scale_colour_manual()

# Using a preset colour package

#install.packages('viridis') #package "viridis"
#install.packages('hexbin')
#library(viridis)
#library(hexbin)

df <- tibble( #fake dataset to create a plot 
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + #new geom
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

## 3.7 Themes

```{r, echo=T, results='hide', fig.show='hide'}
#Trying different default themes
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw() #theme_bw()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light() #theme_light()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic() #theme_classic()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark() #theme_dark()

#Developing a custom theme
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="pink"),
        axis.text.x=element_text(colour="pink"),
        text=element_text(family="Arial"))
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme(panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="pink"),
        axis.text.x=element_text(colour="pink"),
        text=element_text(family="Arial"))
```

## 3.8 Saving and exporting plots

```{r, echo=T, results='hide', fig.show='hide'}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")

```
